---
title: "144L_Abundance_2021_Data"
author: "Gabrielle Mousseau"
date: "10/18/2021"
output: github_document
---

#Goal 
This document sows how individual bottle bacterial abundance from 2021 remineralization experiments were processed, QC'd, and analyzed. Info on how to code and notes are also included. 


#Intro to R Markdown 
To create a new code chunk: Cmd + option + i.
Load packages needed to analyze the data.
Using settings, we can toggle on/off warnings that we dont want in our final rmd.
```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(lubridate)
library(ggplot2)
```


#Import Data
Attach data and join datasets usings variables that are the same across two dataframes
```{r}
excel_sheets("~/144l_students_2021/Input_Data/week4/144L_2021_BactAbund.xlsx")

metadata <- read_excel("~/144l_students_2021/Input_Data/week4/144L_2021_BactAbund.xlsx", sheet = "Metadata")

glimpse(metadata)
#unique(metadata$Bottle)
#unique(metadata$Treatment)

data <- read_excel("~/144l_students_2021/Input_Data/week4/144L_2021_BactAbund.xlsx", sheet = "Data")
glimpse(data)

joined <- left_join(metadata, data) #attach data to metadata

glimpse(joined)
```


#Prepare Data 
We convert categorical data (date and time) to numeric, add columns with time elapsed for each treatment, and concert cells_uL to cells_L to help with TOC data later. We will also subset data for variables of interest and drop NA values.

We will utilize **piping**. Piping allows you to five multiple commands. %>% means "first do this. THEN, do this" Then = %>% giving order to the commands. 
```{r warning=FALSE}
cells <- joined %>% 
  mutate(Datetime = ymd_hm(Datetime),#splits apart Datetime as specified
  cells_L = as.numeric(all_cells_uL) * 1000000) %>% 
  group_by(Treatment, Bottle) %>% 
  #group datasets so we can time elpase properly
  mutate(interv = interval(first(Datetime), Datetime), 
  s = as.numeric(interv), 
  hours = s/3600, 
  days = hours/24) %>% 
  ungroup() %>% 
  select(Experiment:DNA_Sample, cells_L, hours, days) %>%
  drop_na(cells_L)
  
glimpse(cells)
#View(cells)
```

#Plot Growth Curve
```{r}
custom.colors <- c("Control" = "#377E88", "Kelp Exudate" = "#4DAF4A", "Kelp Exudate_Nitrate_Phosphate" = "#E41A1C", "Glucose_Nitrate_Phosphate_" = "#FF7F00")

levels <- c("Control", "Kelp Exudate", "Kelp Exudate_Nitrate_Phosphate", "Glucose_Nitrate_Phosphate")

cells %>% 
  mutate(dna = ifelse(DNA_Sample == T, "*", NA)) %>%
  ggplot(aes(x=days, y = cells_L, group = interaction(Treatment, Bottle))) + 
  geom_line(aes(color = factor(Treatment, levels= levels)), size =1) +
  geom_point(aes(fill = factor(Treatment, levels = levels)),
  size = 3, color = "black", shape = 21) +
  geom_text(aes(label = dna), size = 12, color = "#E41A1C") +
  labs(x = "Days", y = expression(paste("Cells, L"^-1)), fill = "") +
  guides(color = "none") + 
  scale_color_manual(values = custom.colors) + 
  scale_fill_manual(values = custom.colors)
  #facet_grid(rows = "Treatment")
  theme_bw()
  glimpse(cells)
```


#Next Steps
We can calculate:
-total change in cells from initial conditions to the end of the experiment
-specific growth rate as the slope of ln(abundance) vs. time during exponential growth phase
-doubling time as ln(2) divided by the specific growth rate
-mean of each of these parameters across each treatment

First, we need to determine where the exponential growth is occuring in each of our bottles, if it does. To do this, we'll plot ln(abundance) vs. time.

#Identify exponential phase of growth in oiur remin experiments

***Note" logs in R***

log(x) gives the natural log of x, not the log base qo
log10(x) would give the log base of 10
log2(x) gives log base 2
```{r}
ln_cells <- cells %>%
  group_by(Treatment, Bottle) %>%
  mutate(ln_cells = log(cells_L), 
         diff_ln_cells = ln_cells - lag(ln_cells, default = first(ln_cells)))
#View(ln_cells)
```

Now, we can plot our new data!

```{r fig.height=6, fig.width=8}
ln_cells %>%
  mutate(dna = ifelse(DNA_Sample == T, "*", NA)) %>%
  ggplot(aes(x=days, y = diff_ln_cells, group = interaction(Treatment, Bottle))) + 
  geom_line(aes(color = factor(Treatment, levels= levels)), size =1) +
  geom_point(aes(fill = factor(Treatment, levels = levels)),
  size = 3, color = "black", shape = 21) +
  geom_text(aes(label = dna), size = 12, color = "#E41A1C") + 
  labs(x = "Days", y = expression(paste("∆ln cells, L"^-1)),fill = "") +
  guides(color = "none") + 
  scale_color_manual(values = custom.colors) + 
  scale_fill_manual(values = custom.colors) +
  facet_wrap("Bottle", ncol =2) + 
  theme_bw()
```
Exponential growth seems to be occurring right in the beginning of the experiment and again at the end.

What happens if we plot ln_cells? Can we see more growth in the control?

```{r fig.height=6, fig.width=8}
ln_cells %>%
  mutate(dna = ifelse(DNA_Sample == T, "*", NA)) %>%
  ggplot(aes(x=days, y = ln_cells, group = interaction(Treatment, Bottle))) + 
  geom_line(aes(color = factor(Treatment, levels= levels)), size =1) +
  geom_point(aes(fill = factor(Treatment, levels = levels)),
  size = 3, color = "black", shape = 21) +
  geom_text(aes(label = dna), size = 12, color = "#E41A1C") + 
  labs(x = "Days", y = expression(paste("ln cells, L"^-1)),fill = "") +
  guides(color = "none") + 
  scale_color_manual(values = custom.colors) + 
  scale_fill_manual(values = custom.colors) +
  facet_wrap("Bottle", ncol =2) + 
  theme_bw()
```
This shows more growth in the beginning compared to the end as expected!
